---
title: "admixture"
author: "Estelle Chen"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include = FALSE, message=FALSE}
# Modify this setup code chunk to set options
# or add extra packages etc if needed.
# See the project instructions for more details
# on what code to show, and where/how.

# Set default code chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE
)

library(ggplot2)
library(patchwork)
library(magrittr)
theme_set(theme_bw())
library(readxl)
library(patchwork)
suppressPackageStartupMessages(library(tidyverse))
library(MASS)
library(mvtnorm)
library(MCMCpack)
library(coda)
library(DirichletReg)
library(actuar)
library(cubature)

# To give the same random number sequence every time the document is knit:ed,
# making it easier to discuss the specific numbers in the text:
set.seed(12345L)
```

# Data

```{r data, warning=FALSE, message=FALSE}
# read original data
df <- read_excel("Project data.xls")
colnames(df)[1] = "Region"
regions <- df$Region

# replace " ", "/" and "-" with "_" in row and column names
names(df) = gsub("[ , /, -]", "_", names(df))

# African data
afr <- df[5:10, ]
R <- 6
C <- 15
y_rc <- afr[, 2:16]
N_r <- afr[, 17]
# frequency
f_rc <- y_rc 
for (c in 1:C) {
  f_rc[, c] <- f_rc[, c] / N_r
}

# American data
ame <- df[1:4, ]
S <- 4
y_sc <- ame[, 2:16]
N_s <- ame[, 17]
```

# Model likelihood
```{r model}
# probability parameter vector for multinomial distribution in terms of alpha
p_s_multinom <- function(alpha, s) {
  p_s <- colSums(alpha*f_rc)
  p_s
}

# pmf of y_s ~ multinomial(N_s, p_s)
model <- function(alpha, s) {
  y_s <- y_sc[s, ]
  N <- N_s[s, 1]
  p <- p_s_multinom(alpha, s)
  
  dmultinom(y_s, size = N, prob = p)
}
```

# Prior and posterior of alpha

```{r bayesian alpha, message=FALSE}
# prior of alpha ~ Dir(1)
prior_alpha <- function(alpha) {
  ddirichlet(matrix(alpha, nrow = 1), numeric(R)+1)
}

# posterior of alpha
posterior_alpha <- function(alpha, s) {
  model(alpha, s) * prior_alpha(alpha)
}
```

# MH algorithm for alpha

```{r MH alpha}
# define a function to sample alpha using MH algorithm
MHalpha <- function(alpha_init, s, sigma2, I) {
  #' inputs
  #' alpha_init: initial alpha vector (length = R)
  #'             last element = 1 - sum of first R-1 elements
  #' s: which American region
  #' sigma2: variance for the first R-1 elements of candidate alpha*
  #'         covariance matrix for multivariate Normal proposal is sigma2*I_(R-1)
  #' I: number of iterations
  
  # 1. select initial value
  alpha_now <- alpha_init
  R = length(alpha_init)  # dimension
  alpha_out <- matrix(0, nrow = I, ncol = R)  # matrix to store all alpha samples
  accept_num <- 0  # number of accepted candidates
  neg_alpha <- 0  # number of negative alpha's not meeting constraints
  
  
  # 2. for i = 1, ..., I, repeat:
  for (i in 1:I) {
    # a. draw a candidate from proposal
    alpha_cand_before <- rmvnorm(n = 1,
                                 mean = alpha_now[1:R-1], 
                                 sigma = diag(sigma2, nrow = R-1))
    alpha_cand_R <- 1 - sum(alpha_cand_before)
    alpha_cand <- append(alpha_cand_before, c(alpha_cand_R))
    
    if (all(alpha_cand >= 0)) {  # constraints satisfied
      # b. acceptance probability
      # posterior of alpha^(i-1)
      posterior_now <- posterior_alpha(alpha_now, s)
      # posterior of candidate alpha*
      posterior_cand <- posterior_alpha(alpha_cand, s)
      # acceptance probability
      P_accept <- min(posterior_cand / posterior_now, 1)

      # c. generate u from U[0,1]
      u <- runif(1)
      
      # d. accept / reject the candidate
      if (u <= P_accept) {  
        # if u < P_accept, accept alpha_cand and save alpha^i = alpha_cand
        # else, reject alpha_cand and save alpha_now
        alpha_now <- alpha_cand
        accept_num <- accept_num + 1
        posterior_now <- posterior_cand
      } 
    } else {  # constraints not met, reject
      neg_alpha <- neg_alpha + 1
    }
    alpha_out[i, ] <- alpha_now  # store alpha for iteration i  
  }
  
  #' return: 
  #' matrix of alpha
  #' acceptance rate over I iterations 
  #' rate of negative alpha's over I iterations
  list(alpha = alpha_out, accept_rate = accept_num/I, neg_alpha_rate = neg_alpha/I)
}
```

# M-H for z

## Prior and posterior of z

```{r bayesian z}
# prior of z, with each z_rs ~ Gamma(1,1) (likelihood)
prior_z <- function(z) {
  prod(dgamma(z, shape = 1, rate = 1))
}

# posterior of z
posterior_z <- function(z, s) {
  alpha <- z/sum(z)
  model(alpha, s) * prior_z(z)
}
```

## Random walk MH algorithm for z

```{r MH z}
# define a function to sample z using RW MH algorithm
MHz <- function(z_init, s, sigma2, I) {
  #' inputs
  #' z_init: initial z vector (length = R)
  #' s: which American region
  #' sigma2: variance for the elements of candidate z*
  #'         covariance matrix for multivariate Normal proposal is sigma2*I_R
  #' I: number of iterations
  
  # 1. select initial value
  z_now <- z_init
  R = length(z_init)  # dimension
  # I*R matrix to store all z's obtained in I iterations
  z_out <- matrix(0, nrow = I, ncol = R)
  # I*R matrix to store corresponding alpha's
  alpha_out <- matrix(0, nrow = I, ncol = R)
  accept_num <- 0  # number of accepted candidates
  nonpos_z <- 0  # number of non-positive z's not meeting constraints
  
  
  # 2. for i = 1, ..., I, repeat:
  for (i in 1:I) {
    # a. draw a candidate from proposal
    z_cand <- rmvnorm(n = 1, mean = z_now, sigma = diag(sigma2, nrow = R))

    if (all(z_cand > 0)) {  # constraint satisfied
      # b. acceptance probability
      # posterior of z^(i-1)
      posterior_now <- posterior_z(z_now, s)
      # posterior of candidate z*
      posterior_cand <- posterior_z(z_cand, s)
      # acceptance probability
      P_accept <- min(posterior_cand / posterior_now, 1)

      # c. generate u from U[0,1]
      u <- runif(1)
      
      # d. accept / reject the candidate
      if (u <= P_accept) {  
        # if u <= P_accept, accept z_cand and save z^i = z_cand
        # if u > P_accept, reject z_cand and save z_now
        z_now <- z_cand
        accept_num <- accept_num + 1
        posterior_now <- posterior_cand
      } 
    } else {  # constraint not met, reject
      nonpos_z <- nonpos_z + 1
    }
    z_out[i, ] <- z_now  # store z for iteration i
    alpha_out[i, ] <- z_now / sum(z_now)  # calculate and store alpha
  }
  
  #' return: 
  #' matrix of z
  #' matrix of alpha
  #' acceptance rate over I iterations 
  #' rate of non-positive z's over I iterations
  list(z = z_out, alpha = alpha_out, accept_rate = accept_num/I, nonpos_z_rate = nonpos_z/I)
}
```

# MH for x

## Prior and posterior of x

```{r bayesian x}
# prior of x, with each x_rs ~ log(z_rs) (likelihood)
prior_x <- function(x) {
  exp(sum(x - exp(x)))
}

# posterior of x
posterior_x <- function(x, s) {
  alpha <- exp(x)/sum(exp(x))
  model(alpha, s) * prior_x(x)
}
```

## Random walk MH algorithm for x

```{r MH x}
# define a function to sample x using RW M-H algorithm
MHx <- function(x_init, s, sigma2, I) {
  #' inputs
  #' x_init: initial x vector (length = R)
  #' s: which American region
  #' sigma2: variance for the elements of candidate x*
  #'         covariance matrix for multivariate Normal proposal is sigma2*I_R
  #' I: number of iterations
  
  # 1. select initial value
  x_now <- x_init
  R = length(x_init)  # dimension
  # I*R matrix to store all x's obtained in I iterations
  x_out <- matrix(0, nrow = I, ncol = R)
  # I*R matrix to store corresponding alpha's
  alpha_out <- matrix(0, nrow = I, ncol = R)
  accept_num <- 0  # number of accepted candidates
  
  
  # 2. for i = 1, ..., I, repeat:
  for (i in 1:I) {
    # a. draw a candidate from proposal
    x_cand <- rmvnorm(n = 1, mean = x_now, sigma = diag(sigma2, nrow = R))

    # the only constraint is x being real
    # b. acceptance probability
    # posterior of x^(i-1)
    posterior_now <- posterior_x(x_now, s)
    # posterior of candidate x*
    posterior_cand <- posterior_x(x_cand, s)
    # acceptance probability
    P_accept <- min(posterior_cand / posterior_now, 1)

    # c. generate u from U[0,1]
    u <- runif(1)
    
    # d. accept / reject the candidate
    if (u <= P_accept) {  
      # if u <= P_accept, accept x_cand and save x^i = x_cand
      # if u > P_accept, reject x_cand and save x_now
      x_now <- x_cand
      accept_num <- accept_num + 1
      posterior_now <- posterior_cand
    }
    x_out[i, ] <- x_now  # store x for iteration i
    alpha_out[i, ] <- exp(x_now)/sum(exp(x_now))  # calculate and store alpha
  } 
  
  #' return: 
  #' matrix of x
  #' matrix of alpha
  #' acceptance rate over I iterations 
  list(x = x_out, alpha = alpha_out, accept_rate = accept_num/I)
}
```

# Rates of MHalpha, MHz, MHx for comparison

```{r rates MH alpha, warning=FALSE}
alpha_0 <- rdirichlet(1, numeric(R)+1)  # randomly generate initial alpha
sigma2s <- c(4, 1, 0.1, 0.01, 0.001, 0.0001)  # values of sigma2

# dataframe to store the results
rates_MHalpha <- data.frame(sigma2 = c(),
                            accpet_rate = c(),
                            neg_alpha_rate = c())

# conduct MHa and store the results
for (i in 1:length(sigma2s)) {
  sigma2 <- sigma2s[i]
  output_alpha <- MHalpha(alpha_init = alpha_0, s = 1, sigma2, I = 5000)
  
  rates_store <- data.frame(sigma2 = c(sigma2), 
                            accept_rate = c(output_alpha$accept_rate), 
                            neg_alpha_rate = c(output_alpha$neg_alpha_rate))
  
  rates_MHalpha <- rbind(rates_MHalpha, rates_store)
}

rates_MHalpha
```

```{r rates MH z, warning=FALSE}
z_0 <- rgamma(R, shape = 1, rate = 1)  # randomly generate initial z
sigma2s <- c(4, 1, 0.1, 0.01, 0.001, 0.0001)  # values of sigma2

# dataframe to store the results
rates_MHz <- data.frame(sigma2 = c(),
                        accpet_rate = c(),
                        nonpos_z_rate = c())

# conduct MHz and store the results
for (i in 1:length(sigma2s)) {
  sigma2 <- sigma2s[i]
  output_z <- MHz(z_init = z_0, s = 1, sigma2, I = 5000)
  
  rates_store <- data.frame(sigma2 = c(sigma2), 
                            accept_rate = c(output_z$accept_rate), 
                            nonpos_z_rate = c(output_z$nonpos_z_rate))
  
  rates_MHz <- rbind(rates_MHz, rates_store)
}

rates_MHz
```

```{r rates MH x, warning=FALSE}
x_0 <- log(rgamma(R, shape = 1, rate = 1))  # randomly generate initial x
sigma2s <- c(4, 1, 0.1, 0.01, 0.001, 0.0001)  # values of sigma2

# dataframe to store the results
rates_MHx <- data.frame(sigma2 = c(),
                        accpet_rate = c())

# conduct MHz and store the results
for (i in 1:length(sigma2s)) {
  sigma2 <- sigma2s[i]
  output_x <- MHx(x_init = x_0, s = 1, sigma2, I = 5000)
  assign(paste("output_x_", i, sep = ""), output_x)  # assign a name to output
  
  rates_store <- data.frame(sigma2 = c(sigma2), 
                            accept_rate = c(output_x$accept_rate))
  
  rates_MHx <- rbind(rates_MHx, rates_store)
}

rates_MHx
```

# Optimisation of MH
## Sample ranges

```{r x ambitious moves}
# traceplots of samples of x_11
# sigma2 = 4
x11_output_1 = mcmc(output_x_1$x[, 1])
traceplot(x11_output_1, ylab = "x_11", mgp = c(2,1,0))
# sigma2 = 1
x11_output_2 = mcmc(output_x_2$x[, 1])
traceplot(x11_output_2, ylab = "x_11", mgp = c(2,1,0))
# sigma2 = 0.01
x11_output_4 = mcmc(output_x_4$x[, 1])
traceplot(x11_output_4, ylab = "x_11", mgp = c(2,1,0))
# sigma2 = 0.0001
x11_output_6 = mcmc(output_x_6$x[, 1])
traceplot(x11_output_6, ylab = "x_11", mgp = c(2,1,0))
```

# Adaptive Metropolis algorithm for x

```{r AM x}
# define a function to sample x using adaptive MH algorithm
AMx <- function(x_init, s, I, beta) {
  #' inputs
  #' x_init: initial x vector (length = R)
  #' s: which American region
  #' I: number of iterations
  #' beta: in proposal for i>2R
  
  # 1. select initial value
  x_now <- x_init
  R = length(x_init)  # dimension
  # I*R matrix to store all x's obtained in I iterations
  x_out <- matrix(0, nrow = I, ncol = R)
  # I*R matrix to store corresponding alpha's
  alpha_out <- matrix(0, nrow = I, ncol = R)
  accept_num <- 0  # number of accepted candidates
  
  
  # 2. for i = 1, ..., I, repeat:
  for (i in 1:I) {
    if (i <= 2*R) {  # i <= 2R
      # a. draw candidate
      Sigma_0 <- diag(0.1^2/R, nrow = R)  # calculate the covariance matrix 0.1^2*I_R/R 
      x_cand <- rmvnorm(n = 1, mean = x_now, sigma = Sigma_0)
      
      # b. acceptance probability
      posterior_now <- posterior_x(x_now, s)
      posterior_cand <- posterior_x(x_cand, s)
      q_now <- dmvnorm(x_now, mean = x_cand, sigma = diag(0.1^2/R, nrow = R))
      q_cand <- dmvnorm(x_cand, mean = x_now, sigma = diag(0.1^2/R, nrow = R))
      P_accept <- min(posterior_cand / posterior_now, 1)
      
      # c. generate u from U[0,1]
      u <- runif(1)
      
      # d. accept / reject the candidate
      if (u <= P_accept) {  
        x_now <- x_cand
        accept_num <- accept_num + 1
        posterior_now <- posterior_cand
      }
    } else {  # i > 2R
      # a. draw candidate
      Sigma_0 <- diag(0.1^2/R, nrow = R)  # covariance matrix 0.1^2*I_R/R 
      x_samples_now <- rbind(x_init, x_out[1:i-1, ])  # samples x^0, ..., x^(i-1)
      Sigma_i <- cov(x_samples_now)  # empirical covariance matrix
      x_cand <- (1-beta) * rmvnorm(n = 1, mean = x_now, sigma = 2.38^2/R*Sigma_i) + 
                beta * rmvnorm(n = 1, mean = x_now, sigma = Sigma_0)
      
      # b. acceptance probability
      posterior_now <- posterior_x(x_now, s)
      posterior_cand <- posterior_x(x_cand, s)
      #q_now <- (1-beta) * dmvnorm(x_now, mean = x_cand, sigma = 2.38^2/R*Sigma_i) + 
      #         beta * dmvnorm(x_now, mean = x_cand, sigma = Sigma_0)
      #q_cand <- (1-beta) * dmvnorm(x_cand, mean = x_now, sigma = 2.38^2/R*Sigma_i) + 
      #          beta * dmvnorm(x_cand, mean = x_now, sigma = Sigma_0)
      #P_accept <- min((posterior_cand*q_now) / (posterior_now*q_cand), 1)
      P_accept <- min(posterior_cand / posterior_now, 1)
      
      # c. generate u from U[0,1]
      u <- runif(1)
      
      # d. accept / reject the candidate
      if (u <= P_accept) {  
        x_now <- x_cand
        accept_num <- accept_num + 1
        posterior_now <- posterior_cand
      }
    }
    x_out[i, ] <- x_now  # store x for iteration i
    alpha_out[i, ] <- exp(x_now)/sum(exp(x_now))  # calculate and store alpha
  } 
  
  #' return: 
  #' matrix of x
  #' matrix of alpha
  #' acceptance rate over I iterations 
  list(x = x_out, alpha = alpha_out, accept_rate = accept_num/I)
}
```

## Acceptance rates for each s

```{r AM rates, warning=FALSE}
x_0 <- log(rgamma(R, shape = 1, rate = 1))  # randomly generate initial x

# apply AM for x for all s
adapt_out_s1 <- AMx(x_init = x_0, s = 1, I = 10000, beta = 0.01)
adapt_out_s2 <- AMx(x_init = x_0, s = 2, I = 10000, beta = 0.01)
adapt_out_s3 <- AMx(x_init = x_0, s = 3, I = 10000, beta = 0.01)
adapt_out_s4 <- AMx(x_init = x_0, s = 4, I = 10000, beta = 0.01)

# dataframe of acceptance rates
adapt_rates <- data.frame(s = c(1, 2, 3, 4),
                          accept_rate = c(adapt_out_s1$accept_rate, adapt_out_s2$accept_rate,
                                          adapt_out_s3$accept_rate, adapt_out_s4$accept_rate))

adapt_rates
```

## Trace plots for each alpha_rs

```{r AM alpha}
alpha_s1 <- mcmc(adapt_out_s1$alpha)
alpha_s2 <- mcmc(adapt_out_s2$alpha)
alpha_s3 <- mcmc(adapt_out_s3$alpha)
alpha_s4 <- mcmc(adapt_out_s4$alpha)
```

```{r alpha traceplot s=1}
# s=1
for (i in 1:6) {
  traceplot(alpha_s1[, i], ylab = c(paste("alpha_", i, "1", sep = "")), mgp = c(2,1,0))
}
```

```{r alpha traceplot s=2}
# s=2
for (i in 1:6) {
  traceplot(alpha_s2[, i], ylab = c(paste("alpha_", i, "2", sep = "")), mgp = c(2,1,0))
}
```

```{r alpha traceplot s=3}
# s=3
for (i in 1:6) {
  traceplot(alpha_s3[, i], ylab = c(paste("alpha_", i, "3", sep = "")), mgp = c(2,1,0))
}
```

```{r alpha traceplot s=4}
# s=4
for (i in 1:6) {
  traceplot(alpha_s4[, i], ylab = c(paste("alpha_", i, "4", sep = "")), mgp = c(2,1,0))
}
```

## Sample mean and SD for each burnt alpha_rs

```{r AM alpha burnt summary}
alpha_burnt_s1 <- mcmc(alpha_s1[2001:10000, ])
alpha_burnt_s2 <- mcmc(alpha_s2[2001:10000, ])
alpha_burnt_s3 <- mcmc(alpha_s3[2001:10000, ])
alpha_burnt_s4 <- mcmc(alpha_s4[2001:10000, ])

print("North America (s = 1)")
summary(alpha_burnt_s1)

print("Central America (s = 2)")
summary(alpha_burnt_s2)

print("South America (s = 3)")
summary(alpha_burnt_s3)

print("Americas (s = 4)")
summary(alpha_burnt_s4)
```

## Probability of inferred orders of alpha

```{r probability}
# probability of alpha2s being max for posterior samples
prob_alpha2s_max <- function(alpha) {
  I <- dim(alpha)[1]  # number of samples
  N <- 0  # number of samples satisfying the results
  for (i in 1:I) {
    if (max(alpha[i, ]) == alpha[i, 2]) {
      N <- N+1
    }
  }
  N/I
}

# probability of alpha3s being max for posterior samples
# for s=3
prob_alpha3s_max <- function(alpha) {
  I <- dim(alpha)[1]  # number of samples
  N <- 0  # number of samples satisfying the results
  for (i in 1:I) {
    if (max(alpha[i, ]) == alpha[i, 3]) {
      N <- N+1
    }
  }
  N/I
}

# probability of alpha1s < alpha6s < alpha4s < alpha5s for posterior samples
prob_other_alpha <- function(alpha) {
  I <- dim(alpha)[1]  # number of samples
  N <- 0  # number of samples satisfying the results
  for (i in 1:I) {
    if (alpha[i, 1] < alpha[i, 6] & alpha[i, 6] < alpha[i, 4] & alpha[i, 4] < alpha[i, 5]) {
      N <- N+1
    }
  }
  N/I
}

alpha2s_max_probs <- data.frame(s = c(1, 2, 3, 4),
                                probability_alpha2s_max = c(prob_alpha2s_max(alpha_burnt_s1),
                                                            prob_alpha2s_max(alpha_burnt_s2),
                                                            prob_alpha2s_max(alpha_burnt_s3),
                                                            prob_alpha2s_max(alpha_burnt_s4)))
alpha3s_max_probs <- data.frame(s = c(1, 2, 3, 4),
                                probability_alpha3s_max = c(prob_alpha3s_max(alpha_burnt_s1),
                                                            prob_alpha3s_max(alpha_burnt_s2),
                                                            prob_alpha3s_max(alpha_burnt_s3),
                                                            prob_alpha3s_max(alpha_burnt_s4)))
other_alpha_probs <- data.frame(s = c(1, 2, 3, 4),
                                probability = c(prob_other_alpha(alpha_burnt_s1),
                                                prob_other_alpha(alpha_burnt_s2),
                                                prob_other_alpha(alpha_burnt_s3),
                                                prob_other_alpha(alpha_burnt_s4)))
alpha2s_max_probs
alpha3s_max_probs
other_alpha_probs
```

## Trace plots and density plots for each burnt alpha_rs

```{r burnt plot s=1}
for (i in 1:6) {
  traceplot(alpha_burnt_s1[, i], ylab = c(paste("alpha_", i, "1", sep = "")), mgp = c(2,1,0))
  densplot(alpha_burnt_s1[, i], xlab = c(paste("alpha_", i, "1", sep = "")), 
           ylab = "Density", mgp = c(2,1,0))
}
```

```{r burnt plot s=2}
for (i in 1:6) {
  traceplot(alpha_burnt_s2[, i], ylab = c(paste("alpha_", i, "2", sep = "")), mgp = c(2,1,0))
  densplot(alpha_burnt_s2[, i], xlab = c(paste("alpha_", i, "2", sep = "")), 
           ylab = "Density", mgp = c(2,1,0))
}
```

```{r burnt plot s=3}
for (i in 1:6) {
  traceplot(alpha_burnt_s3[, i], ylab = c(paste("alpha_", i, "3", sep = "")), mgp = c(2,1,0))
  densplot(alpha_burnt_s3[, i], xlab = c(paste("alpha_", i, "3", sep = "")), 
           ylab = "Density", mgp = c(2,1,0))
}
```

```{r burnt plot s=4}
for (i in 1:6) {
  traceplot(alpha_burnt_s4[, i], ylab = c(paste("alpha_", i, "4", sep = "")), mgp = c(2,1,0))
  densplot(alpha_burnt_s4[, i], xlab = c(paste("alpha_", i, "4", sep = "")), 
           ylab = "Density", mgp = c(2,1,0))
}
```

## Effective sample size

```{r effective size}
print("s = 1")
effectiveSize(alpha_s1)

print("s = 2")
effectiveSize(alpha_s2)

print("s = 3")
effectiveSize(alpha_s3)

print("s = 4")
effectiveSize(alpha_s4)
```

